Kindly Bent to Free Us

### 2019-07-03 kindly/motivation.tex
motivation

A classical example of linearity at work are to control the uses of resources such
as files or network handles.
%
Session types, notably, allow to statically type check protocols.
For instance, the type \lstinline{!int.!int.?int.end} represent a session
where the program must send two integers, receive an integer, then end the session.
%
In this context, it is essential that channels
are used linearly, since every use of the channel
will make the protocol progress and change the type of the channel.

Session types are usually defined with the following operators:

\begin{center}
  \begin{tabular}{r l}
    $\tau ! \mathcal S$ & Send a value of type $\tau$ then continue with protocol $S$.\\
    $\tau ? \mathcal S$& Receive a value of type $\tau$ then continue with protocol $S$.\\
    $\mathcal S \oplus \mathcal S'$& Choose between protocols $S$ and $S'$.\\
    $\mathcal S \operatorname{\&} \mathcal S'$& Offer a choice between protocols $S$ and $S'$.
  \end{tabular}
\end{center}

It is well known that binary session types can be embedded into ML languages
\cite{DBLP:journals/jfp/Padovani17}. We can easily
embed session types in our language and provide the following API:
%
\begin{lstlisting}
type 'S st : lin
val receive: ('a ? 'S) st -> 'a * 'S st
val send : ('a:'k). 'a -> ('a ! 'S) st -{'k}> 'S st
val create : unit -> 'S st * (dual 'S) st
val close : end st -> unit
\end{lstlisting}

>>>>

Session typing \cite{Honda1993,DBLP:conf/esop/HondaVK98} is a type
discipline for checking protocols statically. A session type is
ascribed to a communication channel and describes
a sequence of interactions. For instance, the type
\lstinline{!int.!int.?int.end} specifies a protocol
where the program must send two integers, receive an integer, and then
close the channel. 
%
In this context, channels must be used linearly, because every use of
a channel ``consumes'' one interaction and hence changes the type of the
channel. That is, after sending two integers on the channel,
the remaining channel has type \lstinline{?int.end}.

Here are some standard type operators for session types \SType:
\begin{center}
  $\SType ::=$
  \begin{tabular}[t]{rl}
    $\tau ! \SType$ & Send a value of type $\tau$ then continue with protocol $\SType$.\\
    $\tau ? \SType$& Receive a value of type $\tau$ then continue with protocol $\SType$.\\
    $\SType \oplus \SType'$& Internal choice between protocols $\SType$ and $\SType'$.\\
    $\SType \operatorname{\&} \SType'$
                    & Offer a choice between protocols $\SType$ and $\SType'$. 
  \end{tabular}
\end{center}

\citet{DBLP:journals/jfp/Padovani17}  has shown how to encode this
style of session typing in ML-like languages, but his implementation
downgrades linearity to a run-time check for affinity. Building on that
encoding we can provide a safe API in \lang that statically enforces
linear handling of channels:
%
\begin{lstlisting}
type 'S st : lin (*@\label{line:session1}*)
val receive: ('a ? 'S) st -> 'a * 'S st (*@\label{line:session2}*)
val send : ('a:'k). 'a -> ('a ! 'S) st -{'k}> 'S st
val send : ('a ! 'S) st -> 'a -{lin}> 'S st
val create : unit -> 'S st * (dual 'S) st
val close : end st -> unit
\end{lstlisting}

Line~\ref{line:session1} introduces a parameterized abstract
type \lstinline{st} which is linear as indicated
by its kind \lstinline{lin}. Its low-level implementation would wrap a
handle for a socket, for example. The \lstinline{receive}  operation
in Line~\ref{line:session2} takes a channel that is ready to receive a
value of type \lstinline{'a} and returns a pair of the value a the
channel at its residual type \lstinline{'S}. It does not matter
whether \lstinline{'a} is restricted to be linear, in fact
\lstinline{receive} is polymorphic in the kind of \lstinline{'a} and
also in the kind of \lstinline{->}. This kind polymorphism is the
default if no constraints are specified.

The \lstinline{send} operation, however,
